---
title: "Hands-on Exercise 4a"
author: "Yi Fang"
date: "May 6, 2025"
date-modified: "last-modified"
format: html
editor: visual
freeze: true
warning: false
---

# Visualising Distribution

## 1 Learning Outcome

Visualising distribution is not new in statistical analysis. Some of the popular statistical graphics methods for visualising distribution were covered in Hands-on Exercise 1. These includes histogram, probability density curve (pdf), boxplot, notch plot and violin plot and how they can be created by using ggplot2. In this chapter, two relatively new statistical graphic methods for visualising distribution will be covered, namely ridgeline plot and raincloud plot by using ggplot2 and its extensions.

## 2 Getting Started

### 2.1 Installing and loading the packages

For the purpose of this exercise, the following R packages will be used:

-   ggridges, a ggplot2 extension specially designed for plotting ridgeline plots,

-   ggdist, a ggplot2 extension spacially desgin for visualising distribution and uncertainty,

-   tidyverse, a family of R packages to meet the modern data science and visual communication needs,

-   ggthemes, a ggplot extension that provides the user additional themes, scales, and geoms for the ggplots package, and

-   colorspace, an R package provides a broad toolbox for selecting individual colors or color palettes, manipulating these colors, and employing them in various kinds of visualisations.

The code chunk below will be used load these R packages into RStudio environment.

```{r}
pacman::p_load(ggdist, ggridges, ggthemes,
               colorspace, tidyverse)
```

### 2.2 Data Import

For the purpose of this exercise, Exam_data.csv will be used.

In the code chunk below, read_csv() of readr package is used to import Exam_data.csv into R and saved it into a tibble data.frame.

```{r}
exam <- read_csv("data/Exam_data.csv")
```

## 3 Visualising Distribution with Ridgeline Plot

[*Ridgeline plot*](https://www.data-to-viz.com/graph/ridgeline.html) (sometimes called *Joyplot*) is a data visualisation technique for revealing the distribution of a numeric value for several groups. Distribution can be represented using histograms or density plots, all aligned to the same horizontal scale and presented with a slight overlap.

Figure below is a ridgelines plot showing the distribution of English score by class.

```{r}
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

::: callout-note
-   Ridgeline plots make sense when the number of group to represent is medium to high, and thus a classic window separation would take to much space. Indeed, the fact that groups overlap each other allows to use space more efficiently. If you have less than 5 groups, dealing with other distribution plots is probably better.

-   It works well when there is a clear pattern in the result, like if there is an obvious ranking in groups. Otherwise group will tend to overlap each other, leading to a messy plot not providing any insight.
:::

### 3.1 Plotting Ridgeline Graph: ggridges method

There are several ways to plot ridgeline plot with R. This section will cover how to plot ridgeline plot using [ggridges](https://wilkelab.org/ggridges/index.html) package.

ggridges package provides two main geom to plot gridgeline plots, they are: [`geom_ridgeline()`](https://wilkelab.org/ggridges/reference/geom_ridgeline.html) and [`geom_density_ridges()`](https://wilkelab.org/ggridges/reference/geom_density_ridges.html). The former takes height values directly to draw the ridgelines, and the latter first estimates data densities and then draws those using ridgelines.

The ridgeline plot below is plotted by using `geom_density_ridges()`.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

## Code

```{r}
#| eval: false
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```
:::

### 3.2 Varying fill colors along the x-axis

To fill the area under the ridgeline with varying colors instead of a single solid color, one can use either `geom_ridgeline_gradient()` or [`geom_density_ridges_gradient()`](https://wilkelab.org/ggridges/reference/geom_ridgeline_gradient.html). Both geoms work just like `geom_ridgeline()` and `geom_density_ridges()`, except that they allow for varying fill colors. However, they do not allow for alpha transparency in the fill. For technical reasons, we can have changing fill colors or transparency but not both.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS,
           fill = stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Temp. [F]",
                       option = "C") +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

## Code

```{r}
#| eval: false
ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS,
           fill = stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Temp. [F]",
                       option = "C") +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```
:::

### 3.3 Mapping the probabilities directly onto color

Beside providing additional geom objects to support the need to plot ridgeline plot, ggridges package also provides a stat function called [`stat_density_ridges()`](https://wilkelab.org/ggridges/reference/stat_density_ridges.html) that replaces [`stat_density()`](https://ggplot2.tidyverse.org/reference/geom_density.html) of ggplot2.

Figure below is plotted by mapping the probabilities calculated by using `stat(ecdf)` which represent the empirical cumulative density function for the distribution of English score.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = 0.5 - abs(0.5-stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient", 
                      calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()
```

## Code

```{r}
#| eval: false
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = 0.5 - abs(0.5-stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient", 
                      calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()
```
:::

::: callout-important
It is important include the argument `calc_ecdf = TRUE` in `stat_density_ridges()`.
:::

### 3.4 Ridgeline plots with quartile lines

The ridgeline plot can be colored based on quartile by using `geom_density_ridges_gradient()`. This is calculated using `stat(quantile)` aesthetic as shown in the figure below.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = 4,
    quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()
```

## Code

```{r}
#| eval: false
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = 4,
    quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()
```
:::

Instead of using number to define the quartiles, we can also specify quartiles by cut points such as 2.5% and 97.5% tails to colour the ridgeline plot as shown in the figure below.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = c(0.025, 0.975)
    ) +
  scale_fill_manual(
    name = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")
  ) +
  theme_ridges()
```

## Code

```{r}
#| eval: false
ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = c(0.025, 0.975)
    ) +
  scale_fill_manual(
    name = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")
  ) +
  theme_ridges()
```
:::

## 4 Visualising Distribution using Raincloud Plot

Raincloud Plot is a data visualisation techniques that produces a half-density to a distribution plot. It gets the name because the density plot is in the shape of a “raincloud”. The raincloud (half-density) plot enhances the traditional box-plot by highlighting multiple modalities (an indicator that groups may exist). The boxplot does not show where densities are clustered, but the raincloud plot does!

In this section, one will learn how to create a raincloud plot to visualise the distribution of English score by race. It will be created by using functions provided by **ggdist** and ggplot2 packages.

### 4.1 Plotting a Half Eye graph

First, plot a Half-Eye graph by using [`stat_halfeye()`](https://mjskay.github.io/ggdist/reference/stat_halfeye.html) of **ggdist** package.

This produces a Half Eye visualization, which is contains a half-density and a slab-interval.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA)
```

## Code

```{r}
#| eval: false
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA)
```
:::

::: {.callout-note .callout-tip title="Things to learn from the code chunk above"}
We remove the slab interval by setting .width = 0 and point_colour = NA. :::
:::

### 4.2 Adding the boxplot with `geom_boxplot()`

Next, add the second geometry layer using [`geom_boxplot()`](https://r4va.netlify.app/chap09) of ggplot2. This produces a narrow boxplot. We reduce the width and adjust the opacity.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA)
```

## Code

```{r}
#| eval: false
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA)

```
:::

### 4.3 Adding the Dot Plots with stat_dots()

Next, dd the third geometry layer using [`stat_dots()`](https://mjskay.github.io/ggdist/reference/stat_dots.html) of ggdist package. This produces a half-dotplot, which is similar to a histogram that indicates the number of samples (number of dots) in each bin. Select side = “left” to place it on the left-hand side.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 2)
```

## Code

```{r}
#| eval: false
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 2)
```
:::

### 4.4 Finishing touch

Lastly, [`coord_flip()`](https://ggplot2.tidyverse.org/reference/coord_flip.html) of ggplot2 package will be used to flip the raincloud chart horizontally to give it the raincloud appearance. At the same time, `theme_economist()` of ggthemes package is used to give the raincloud chart a professional publishing standard look.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 1.5) +
  coord_flip() +
  theme_economist()
```

## Code

```{r}
#| eval: false
ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 1.5) +
  coord_flip() +
  theme_economist()
```
:::

## 5 References

**Reference**

-   [Introducing Ridgeline Plots (formerly Joyplots)](https://blog.revolutionanalytics.com/2017/07/joyplots.html)

-   Claus O. Wilke [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/) especially Chapter [6](https://clauswilke.com/dataviz/visualizing-amounts.html), [7](https://clauswilke.com/dataviz/histograms-density-plots.html), [8](https://clauswilke.com/dataviz/ecdf-qq.html), [9](https://clauswilke.com/dataviz/boxplots-violins.html) and [10](https://clauswilke.com/dataviz/visualizing-proportions.html).

-   Allen M, Poggiali D, Whitaker K et al. [“Raincloud plots: a multi-platform tool for robust data. visualization”](https://wellcomeopenresearch.org/articles/4-63) \[version 2; peer review: 2 approved\]. Welcome Open Res 2021, pp. 4:63.

-   [Dots + interval stats and geoms](https://mjskay.github.io/ggdist/articles/dotsinterval.html)
